The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2020-05-09 response_to_referees.tex

\documentclass[11pt]{article}
\usepackage{eigilscmds}
\usepackage{tikzit}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{csquotes}
\input{comonoids.tikzdefs}
\input{comonoids.tikzstyles}

% draft packages
%\usepackage{todonotes}
%\usepackage{showkeys}

\author{Tobias Fritz and Eigil Fjeldgren Rischel}
\title{Response to Referees:\\ The zero--one laws of Kolmogorov and Hewitt--Savage\\ in categorical probability}
\date{\today}

\renewcommand{\sf}{\mathsf}
\DeclareMathOperator{\del}{del}
\begin{document}
\maketitle

\begin{abstract}
	We would like to thank both referees for their useful suggestions. There are some items that we would like to respond to, in particular from the first referee's report.
\end{abstract}

\section*{Referee 1}

We have been quite surprised by the length and the amount of detail of this report, and it is exciting that the referee has invested so much work into it. Unfortunately we disagree with the referee on most counts.

\todo[inline]{Comment first on the referee's suggestion to change the emphasis to infinite tensor products}

To illustrate the general incoherence of the referee's report, we start by commenting on the summary report and then moving on the material in the attached PDF.

\begin{enumerate}[label=(\alph*).]
	\item The referee writes:
		\begin{displayquote}
			Your definition of the infinite tensor product is not correct - it does not capture the property characterizing the Kolmogorov extension theorem.  This can be corrected easily enough.
		\end{displayquote}
		This seems to claim that there is an error in our paper.
		\begin{displayquote}	
		However, that correction does require additional material proving that the Kolmogorov extension theorem really is equivalent to the existence of infinite tensor products for the category of standard Borel spaces. The difficulty arises because it is necessary to show that the Kleisi category construction, which is the framework in which  the Kolmogorov extension theorem is given, is a special case of the more general situation in a Markov semicartesian symmetric monoidal category where we have an inclusion of a subcategory (specified by the  deterministic morphisms) into a category which consist of deterministic and nondeterministic mappings.
		\end{displayquote}
		Now the error seems to have become a mere gap. Regardless, there is no such error or gap: the connection with the Kolmogorov extension theorem is explained in Remark 3.5 and Example 3.6. In order to see that this recovers the extension theorem, it is not necessary to construct $\BorelStoch$ as the Kleisli category of the Giry monad (although this is certainly one way to do it); the concrete construction of $\BorelStoch$ in terms of Markov kernels and the Chapman--Kolmogorov formula is perfectly enough, and we have explained why countable tensor products in our sense exist in $\BorelStoch$, and it is obvious that this specializes to the Kolmogorov extension theorem for countable products of standard Borel spaces.

		In fact, the referee already seems to agree with this: the relevant diagram is exactly the referee's one on the second half of p.6 (in the report). The ``new terminology'' that the referee proposes surely does not have any bearing on whether the universal property described by the diagram holds in $\BorelStoch$, right?
\end{enumerate}

On the comments in the PDF:

\begin{enumerate}[resume,label=(\alph*).]
	\item The referee's proposed formulation of Definition 3.1 is along the lines of our Definition 4.1.\footnote{With the only difference between the preservation by the functors $-\otimes Y$, which we get to in the next item.} Both require the Kolmogorov product to the limit of the finite products $\bigotimes_{i \in F} X_i$ in $\cC$ and in $\cC_{\det}$. So it is not clear to us what the referee's point is.

		Perhaps the referee's suggestion is to replace Definition 3.1 by Definition 4.1. We do not want to do this, since the latter only applies in Markov categories, while the former more generally makes sense in semicartesian monoidal categories.
		\begin{displayquote}
			and hence every ``infinite tensor product'' is Kolmogorov.
		\end{displayquote}
		This is true in categories like $\Stoch$ and $\BorelStoch$, but not in general. We had already explained this point after Definition 4.1.
	\item Quoting from p.7,
		\begin{displayquote}
			I did not incorporate the property of preservation of lim $\mathcal{D}$ by the endofunctor
			$-\otimes Y : \cC_{\det} \to \cC_{\det}$ because that condition is a global property in the sense
		that $Y$ is an arbitrary element in your space which may have no relationship
		to your fixed family $(X_j)_{j\in J}$ for which you are trying to find an extension.
		\end{displayquote}
		Assuming that ``space'' means category, we agree that there is no relation between $Y$ and the family $(X_i)$, but we do not see why and how this would be an argument against our preservation condition. As the referee will hopefully have noticed, this preservation condition is an important property which we use throughout the paper.

		Perhaps the referee's point is that this preservation condition never appears in the context of Kolmogorov's extension theorem. This is true, and the reason is simple: it \emph{automatically holds} in $\BorelStoch$, as we have explained in Example 3.6. Hence there is no reason to state it explicitly in the standard probability context.
	\item On p.8, the referee again seems to conclude the incorrectness of a statement from their lack of understanding of our arguments:
		\begin{displayquote}
		(2) Page 17, the last sentence. The statement
			\begin{displayquote}
			Moreover, $\hat{\sigma}$ is deterministic by Proposition 4.3.
			\end{displayquote}
		is not correct.
		\end{displayquote}
		The comment that follows:
			\begin{displayquote}
				then please cite a reference	
			\end{displayquote}
		indeed makes it clear that the referee does not claim to have found a mistake. In fact, the reason that $\hat{\sigma}$ is deterministic is by a trivial application of the universal property of the Kolmogorov product as a limit in $\cC_{\det}$, based precisely on the facts that the referee mentions: both the composite and the $\pi_F$'s are deterministic, and hence the factorization exists uniquely.
	\item Again p.8:
		\begin{displayquote}
			(3) Corollary 6.1. (Kolmogorov zero-one law) Your assumption, which says that
			$T \subseteq \Omega$, is incorrect, because if $T$ is just a subset of $\Omega$, then I can just choose
			it to be any nonmeasurable set of $\Omega$.
		\end{displayquote}
		It is a mystery to us how it is possible to read over half of a sentence, where the relevant measurability assumption is stated in clear and explicit terms.
		\begin{displayquote}
			If you
			meant to say that $T$ is a measurable set in $\Omega$, the whole problem statement is an
			elementary result in measure theory, and certainly not the Kolmogorov zero-one
			law.
		\end{displayquote}
		If $T$ was an \emph{arbitrary} measurable set in $\Omega$, then the statement would clearly be the false. In any case, our statement of the Kolmogorov zero--one law has been and is exactly the classical one, and it is so classical that it should not require a reference.
	\item p.9, item (2): Even if most categorical foundations in algebra and geometry required monoidal closure (which is obviously not the case, take e.g.~abelian categories), we do not see how this would be an argument for monoidal closure in the context of probability theory.
	\item p.9, item (4): Standard usage of the term ``diagram'' in category theory does not have any association with graph homomorphisms. In standard parlance as we use it, ``diagram'' is just any functor from a small category into the category under consideration.\footnote{See e.g.~\href{https://ncatlab.org/nlab/show/diagram\#CategoryShapedDiagramFunctorially}{https://ncatlab.org/nlab/show/diagram\#CategoryShapedDiagramFunctorially}.}
	\item p.11:
		\begin{displayquote}
			\textbf{Theorem 4.1.} The weak limit of the functor $\mathcal{D}' : \mathscr{J} \to \Stoch$ exists. Moreover, the weak limit of the functor $\mathcal{D}'$ is never a limit in $\Stoch$ except for trivial cases.
		\end{displayquote}
		It is not clear to us what ``the weak limit'' means, since weak limits are rarely unique (up to unique iso), and in particular uniqueness does not hold in the case under consideration. We agree that the referee's construction indeed produces a weak limit that is not a limit. But the fact that that particular weak limit is not a limit says nothing about whether the functor has a limit or not.
\end{enumerate}

\section*{Referee 2}

We thank the referee for a number of suggestions which have led to an improvement of the presentation. It is good to know for us that the paper is not so easy to follow at times. While an exposition of string diagrams and Markov categories would perhaps be a bit too much for this paper, we hope that these shortcomings can be addressed in a possible future textbook treatment of the subject.

%\bibliographystyle{plain}
%\bibliography{categorical-zero-one}

\end{document}

### 2020-05-09 categorical-zero-one.tex
mention unitality
first version of response to referee

This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}). Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry: here, points and lines are described only in terms of their relationships with one another; while in the analytic approach often attributed to Descartes, where geometric objects are defined concretely in terms of more primitive notions like sets and real numbers.

@@

\begin{example}
	\label{stoch}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. A \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a map
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, $f$ assigns to every $x \in X$ a random element of $Y$. Composition is again defined by a variant of the Chapman--Kolmogorov equation, and the symmetric monoidal structure is defined in terms of the usual product of measurable spaces. The monoidal unit $I$ is given by any one-element set with its unique $\sigma$-algebra. As a special case, morphisms $I \to (X,\Sigma_X)$ can be identified with probability measures $\Sigma_X \to [0,1]$.

@@

\begin{example}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[t]$, where the following counterexample similar to~\cite[Example~11.33]{markov_cats} can be formulated.
Consider the additive maps given as the additive extensions of
\begin{align*}
	f(t^{n}) = {}& \begin{cases}t^{n-1} & n \geq 1 \\ 1 & n = 0 \end{cases}\\[3pt]
	g(t^{n}) = {}& \begin{cases}t & n \geq 1 \\ 1 & n= 0\end{cases}\\[3pt]
	h_{1}(t^{n}) = {}& t^{n}\\[3pt]
	h_{2}(t^{n}) = {}& 1
\end{align*}
On the basis monomials $t^n$, we have $(fg)(t^n) = 1$, and therefore
\[
	(fg)(h_1(t^n)t^m) = 1 = (fg)(h_2(t^n)t^m)
\]
for all $n,m \in \mathbb{N}$, so that the hypothesis holds.
If $\CRing_{+}^{\op}$ were causal, then we would also have
\[
	f(g(h_{1}(t^n) t^m) t^\ell) = f(g(h_{2}(t^n) t^m) t^\ell)
\]
for all $n,m,\ell \in \mathbb{N}$. But this clearly fails for $n = \ell = 1$ and $m = 0$.
\end{example}

@@

\begin{problem}
	Is there a Markov category $\cC$ with countable Kolmogorov products such that \Cref{thm:kolmog} specializes to the standard Kolmogorov zero--one law in its general form applicable to all measurable spaces?
  \end{problem}

@@

\section{Speculation}
\todo[inline]{not sure if this section is any good}
It would good to have a way of using this theorem without assuming that the category in question has these infinite tensor products.
We can always make sense of the notion of an infinite independent collection of random variables (just a collection of maps $I \to X_i$).
Of course to talk about an outcome or a function which depends on the outcome of all the $X_i$, we need a map $\prod_i X_i \to T$ of some type.
But since this map is deterministic, we may want to talk about it without admitting it to our category of Markov kernels.

As an example, the probability theory of finite or countable sets is much simpler than the probability theory of infinite sets, since the whole edifice of $\sigma$-algebras can be done away with.
This is a reason to prefer $\sf{FinStoch}$ over the full $\sf{Stoch}$.
In this case, it would be desirable to be able to apply this theorem, given a collection of finite variables $I \to X_i$ and a function $\prod_i X_i \to Y$
(which may be required to satisfy some condition like continuity in the product topology), without begin required to invent the larger category $\sf{Stoch}$.

For instance, maybe there is a Markov category with objects the profinite sets, which can be obtained from $\sf{FinStoch}$ by some formal construction (maybe even simply as the pro-category), which admits infinite tensor products, and where one can profitably interpret this theorem.

>>>>

This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}). Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry: here, points and lines are described only in terms of their relationships with one another; while in the analytic approach often attributed to Descartes, geometric objects are defined concretely in terms of more primitive notions like sets and real numbers.

@@

\begin{example}
	\label{stoch}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. A \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a map
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, one can think of $f$ as assigning to every $x \in X$ a random element of $Y$. Composition is again defined by a variant of the Chapman--Kolmogorov equation, and the symmetric monoidal structure is defined in terms of the usual product of measurable spaces. The monoidal unit $I$ is given by any one-element set with its unique $\sigma$-algebra. As a special case, morphisms $I \to (X,\Sigma_X)$ can be identified with probability measures $\Sigma_X \to [0,1]$.

@@

\begin{example}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[t]$, where the following counterexample similar to~\cite[Example~11.33]{markov_cats} can be formulated.
Consider the additive unital maps given as the additive extensions of
\begin{align*}
	f(t^{n}) = {}& \begin{cases}t^{n-1} & n \geq 1 \\ 1 & n = 0 \end{cases}\\[3pt]
	g(t^{n}) = {}& \begin{cases}t & n \geq 1 \\ 1 & n= 0\end{cases}\\[3pt]
	h_{1}(t^{n}) = {}& t^{n}\\[3pt]
	h_{2}(t^{n}) = {}& 1
\end{align*}
On the basis monomials $t^n$, we have $(fg)(t^n) = 1$, and therefore
\[
	(fg)(h_1(t^n)t^m) = 1 = (fg)(h_2(t^n)t^m)
\]
for all $n,m \in \mathbb{N}$, so that the hypothesis holds.
If $\CRing_{+}^{\op}$ were causal, then we would also have
\[
	f(g(h_{1}(t^n) t^m) t^\ell) = f(g(h_{2}(t^n) t^m) t^\ell)
\]
for all $n,m,\ell \in \mathbb{N}$. But this clearly fails for $n = \ell = 1$ and $m = 0$.
\end{example}

@@

\begin{problem}
	Is there a Markov category $\cC$ with countable Kolmogorov products such that \Cref{thm:kolmog} specializes to the standard Kolmogorov zero--one law, in its general form applicable to all measurable spaces?
  \end{problem}

@@
